Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-08 paper-ieee-trans.tex
cosmetic changes

%\IEEEPARstart{W}hen e-Infrastructures in Europe were first conceived, new
%services to access distributed resources -- data, computing, instrumental --
%were initially considered. Major investments were then injected in order to
%develop distributed solutions that provided a seamless and transparent access
%to those resources, available for the researchers. One of the most known examples is Worldwide
%Large Hadron Collider Computing Grid (WLCG) global computing infrastructure, powered by the
%Particle Physics scientific community, built up to satisfy the computing needs
%of processing, storing and analysing the huge quantity of data coming from the Large Hadron
%Collider particle accelerator.

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
which supports an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One of the well-known examples is the Worldwide 
Large Hadron Collider (LHC) Computing Grid (WLCG) global computing infrastructure, powered 
by the particle physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the Large 
Hadron Collider particle accelerator. A significant number of innovation projects and
technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
and ultimately by the Horizon 2020 programme\footnote{HORIZON 2020 - The EU Framework
Programme for Research and Innovation https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
develop distributed solutions that provide a seamless and transparent access to those
computing and data resources, available for the researchers.

@@

Analyzing the European Commission (EC)-funded software projects in the last decade
we have observed a continuous increase of the prominence and robustness of
testing and validation procedures. One of many reasons is related to the
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology (ICT)
field (such as virtualization, the enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization). All of which
complemented by the emergence of automation and event-response tools.

In this paper, we introduce a set of EC-funded projects
that go from 2001 up to now, in which the authors of this summary were involved.
For each project, we describe the challenges arised, the methodologies used and
the achievements towards software reliability. The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the 
experiences from the recent INDIGO--DataCloud project. Finally, Section \ref{sec:con} concludes.

@@

The DataGrid \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
had as main goal to develop the software to provide basic Grid functionality
and associated management tools for a large scale testbed for demonstration projects in three
specific areas of science, such as particle physics, bioinformatics and earth observation.
%was devoted to develop the first European infrastructure for Grid computing. 
The project brought together 21 academic and industry partners, from 15 different
countries. 
The project delivered its own software distribution, named EDG (EU DataGrid), strongly 
based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
which 5 were devoted to software development coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

@@

The three phases of Enabling Grids for E--sciencE (EGEE, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from more than 240 institutions in 45 countries to
provide a seamless Grid infrastructure for e--Science. EGEE--II and EGEE--III
featured the internationalization of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements from the diverse scientific communities would need to develop a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

The {\sl gLite} middleware \cite{glite} was the ultimate
official software distribution of EGEE as of 2006, after two years of prototyping and
re--engineering efforts to converge with LHC Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software distributions. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, that issued more than 10 thousand bug fixes, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

The source code was available at a private, centralized version control system.
The code was passed through a manual certification procedure at the time of the release. 
This procedure tried to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process 
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
production. Starting with EGEE--II, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the ETICS \cite{etics} solution.

@@

The E--Infrastructure for Testing, Integration and Configuration of Software
\cite{cordis:etics, cordis:etics2} (ETICS, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects such as EGEE and its gLite middleware. The framework
integrated different technologies and tools in order to provide automated configuration,
build and testing capabilities, as well as auto-generated documentation and
software metrics gathering such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics}. The ETICS framework was the first automated
service for delivering quality software products in distributed environments like
the Grids.

@@

The European Middleware Initiative (EMI, May 2010 -- Apr 2013)
\cite{cordis:emi} project joined the 4 major Grid middleware providers in
Europe at the time -- {\sl gLite}, {\sl UNICORE}, {\sl ARC} and {\sl dCache} --
with the goal to maintain and evolve the middleware focusing on extending their
interoperability and improving the reliability of the services. The
ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
characteristics that needed to be present in the EMI software products and
processes to be able to meet the EMI quality requirements
\cite{emi-quality-model}.

For each software characteristic, a set of associated
metrics and Key Performance Indicators (KPIs) were identified and defined in
detail in the EMI Metrics Specification \cite{emi-metrics}. The project
leveraged the {\sl ETICS} service for the development, continuous integration and release management,
as well as for metric tracking, making queries on the collected data to display
them through the chart generation framework.

@@

The EGI--InSPIRE (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
arose as the continuation of the EGEE--III project, with the mission of establishing and
maintaining a sustainable European Grid Infrastructure.

EGI--InSPIRE not only ensured the coordination of the operation of a secure, reliable 
European-wide production Grid infrastructure federated from National Grid Initiatives (NGIs)
that is integrated and interoperates with other Grids worldwide but also foreseed the 
maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
the EGI Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
coordination of the external provision of software required by EGI to 
form the production infrastructure. For the later it maintained a production--ready 
Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution), and 
defined general and component specific quality criteria to be applied to software 
components made available through this repository.
The role of {\sl UMD} was to enforce the fulfilment of this set of quality
criteria \cite{egi-qc} in the software being delivered (i.e.,
{\sl EMI} and {\sl Globus}) \cite{mario}. {\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under a follow--up
project, the EGI--Engage (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The INDIGO--DataCloud (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

INDIGO--DataCloud is the most recent project included in this paper thus it
emerged from the lessons learned in the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of
being framed into large collaborative projects with an extensive list of partners.

@@

%The INDIGO--Datacloud project reflects the progress made in software
%quality and reliability aspects throughout the past scientific European
%experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
%highly influenced by the insights of current big worldwide collaborations of
%software development. These collaborations have continuously inspired the use of new
%technologies and procedures to increase the robustness and quality of the
%software being delivered. In this scenario, the implementation of the DevOps culture 
%is one of the major outcomes, as this paradigm has been progressively adopted by the
%project.

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth]{images/codestyle.png}
\caption{Code style standards followed by INDIGO--DataCloud's software products.}
\label{fig:fig_codestyle}
\end{figure}

@@

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
over the INDIGO--DataCloud software stack.

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major INDIGO--DataCloud releases, denoting an incremental 
trend over the course of the project. As depicted in the Figure, 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

@@

\item \textit{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation, 
self-documenting code, as they sequentially guide the component's deployment process in multiple
platforms. In this regard, INDIGO--DataCloud project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The INDIGO--DataCloud project promoted the application of a CI scenario that enforces,
for any piece of software produced, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as 
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline so to have a per-change trend evolution of the common source code 
related metrics, such as the source lines of code (SLOC) or the cyclomatic complexity.

@@

Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
INDIGO--DataCloud's first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer.

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually more correlated. The software updates of 
INDIGO--DataCloud products, taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI-validated 
software components are automatically delivered in online repositories without the need of 
any human interaction but ultimately requiring expert supervision to validate the results.

INDIGO--DataCloud software stack is delivered in form of Linux software packages (rpms, debs) 
and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
package pipeline} intentionally delivers the packages produced in a pre-production -- preview --
branch. Before that, once the software packages have been created, they are uploaded to a 
testing branch. The available automated deployment solution -- see \cite{subsec:sqa} -- checks
the installation and configuration of the component relying on the testing repository. On 
successful completion, the packages are then automatically signed and subsequently moved to 
the preview branch, from where they finally be made available in the production branch after
the release manager supervises the process.

The \textit{Docker container pipeline} uses instead the automated deployment solution to 
actually install and configure the software component in the container image. Once this is 
done the recently created image is uploaded to the production DockerHub repository, tagged
with the label corresponding to the current INDIGO-DataCloud release. Figure 
\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation, 
automatically triggered by a change in the source code.

@@

\begin{figure*}	
	\centering
	\begin{subfigure}
		\centering
\includegraphics[width=0.85\textwidth]{images/disvis-flow.png}
\caption{DevOps pipeline to distribute Docker images for Disvis application.}
\label{fig:fig_disvis}		
	\end{subfigure}
	\quad
	\begin{subfigure}
		\centering
\includegraphics[width=\textwidth]{images/pilotpreview.png}
\caption{Resource centers supporting the Pilot Preview testbed and corresponding
set of deployed INDIGO--DataCloud components or services.}
\label{fig:fig_pilotpreview}
	\end{subfigure}
\end{figure*}

@@

The experience gathered throughout the project with regard to the adoption of
different DevOps practices is not only useful and suitable for the software related
to the core services in the INDIGO--DataCloud solution, but also applicable to the
development and distribution of the applications coming from the user communities.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the INDIGO--DataCloud’s application repository.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the INDIGO--DataCloud
solutions and use cases validation, by deploying and executing the applications
with the last stable version of the software in environments as similar as 
possible to production ones. A map of the Pilot Preview
infrastructure is depicted in Figure \ref{fig:fig_pilotpreview}, it shows the
resources providers and the components or services they deployed and supported.

>>>>

\IEEEPARstart{D}uring the last 15 years the provision of an European e-Infrastructure,
which supports an unified access to large-scale computing and intense data analysis
applications, has been driven by the needs of scientific communities as part of their
pan-European research collaborations. One of the well-known examples is the Worldwide 
Large Hadron Collider (LHC) Computing Grid infrastructure, powered 
by the particle physics scientific community, built up to satisfy the computing needs
of processing, storing and analysing the huge quantity of data coming from the Large 
Hadron Collider particle accelerator. A significant number of innovation projects and
technology transfer activities -- embraced by the 5th, 6th and 7th Framework Programmes,
and ultimately by the Horizon 2020 programme\footnote{HORIZON 2020 - The EU Framework
Programme for Research and Innovation https://ec.europa.eu/programmes/horizon2020/} -- are being held to coordinate and 
develop distributed solutions that provide a seamless and transparent access to those
computing and data resources, available for the researchers.

@@

Analyzing the European Commission (EC)-funded software projects in the last decade
we have observed a continuous increase of the prominence and robustness of
testing and validation procedures. One of many reasons is related to the
evolution of the software engineering methodologies (creation of the Agile
Manifesto \cite{agile-manifesto}, and rise of DevOps practices \cite{zhu}) together
with the parallel advancements in the information and communication technology
field (such as virtualization, the enabling technology for Cloud computing and 
operating--system--level virtualization known as containerization). All of which
complemented by the emergence of automation and event-response tools.

In this paper, we introduce a set of EC-funded projects
that go from 2001 up to now, in which the authors of this summary were involved.
For each project, we describe the challenges arised, the methodologies used and
the achievements towards software reliability. The reminder of this paper is
organized as follows. Section \ref{sec:ev} details the evolution of research on
software reliability in EC-projects. Section \ref{sec:ntsr} provides
information about new research trends in software reliability based upon the 
experiences from the recent {\sl INDIGO--DataCloud} project. Finally, Section \ref{sec:con} concludes.

@@

The {\sl DataGrid} \cite{cordis:datagrid} project (Jan 2001 -- Dec 2003) 
had as main goal to develop the software to provide basic Grid functionality
and associated management tools for a large scale testbed for demonstration projects in three
specific areas of science, such as particle physics, bioinformatics and earth observation.
%was devoted to develop the first European infrastructure for Grid computing. 
The project brought together 21 academic and industry partners, from 15 different
countries. 
The project delivered its own software distribution, named EDG (EU {\sl DataGrid}), strongly 
based on Globus middleware services \cite{globus}. The project was organized in 12 workpages, from 
which 5 were devoted to software development coordinated by an 
Architecture Task Force, supervising the overall design and technical consistency 
of the developments.

@@

The three phases of Enabling Grids for E--sciencE ({\sl EGEE}, Apr 2004 -- Apr 2010)
\cite{cordis:egee, cordis:egee2, cordis:egee3} projects brought together
scientists and engineers from more than 240 institutions in 45 countries to
provide a seamless Grid infrastructure for e--Science. {\sl EGEE--II} and {\sl EGEE--III}
featured the internationalization of the project, embracing worldwide research
institutions and user communities. The software to sustain the increasing
requirements from the diverse scientific communities would need to develop a
rich set of new services while maintaining a sustainable infrastructure for
Grid computing, eventually used by more than 15 thousand researchers and deployed in
over 250 institutions.

The {\sl gLite} middleware \cite{glite} was the ultimate
official software distribution of {\sl EGEE} as of 2006, after two years of prototyping and
re--engineering efforts to converge with {\sl LHC} Computing Grid (LCG--2), Virtual
Data Toolkit (VDT) and Condor \cite{condor} software distributions. The
development team was comprised of more that 80 people from 12 academic and
industrial partners, that issued more than 10 thousand bug fixes, 1.7 thousand patches and
defined over 300 development tasks tracked by the use of bug/task management tools.

The source code was available at a private, centralized version control system.
The code was passed through a manual certification procedure at the time of the release. 
This procedure tried to improve the reliability of the software components by applying an
acceptance criteria plan at each stage of the software release process 
\cite{egee:acceptance-criteria} i.e. integration, certification, pre--production and 
production. Starting with {\sl EGEE--II}, the project adopted automation in the software
lifecycle process by leveraging the automatic build system for Grid middleware, that is
the {\sl ETICS} \cite{etics} solution.

@@

The E--Infrastructure for Testing, Integration and Configuration of Software
\cite{cordis:etics, cordis:etics2} ({\sl ETICS}, Jan 2006 -- Feb 2010) project aimed
at addressing the challenges in producing quality software in distributed,
collaborative projects such as {\sl EGEE} and its {\sl gLite} middleware. The framework
integrated different technologies and tools in order to provide automated configuration,
build and testing capabilities, as well as auto-generated documentation and
software metrics gathering such as Source Lines Of Code (SLOC), complexity and
number of defects/bugs \cite{etics}. The {\sl ETICS} framework was the first automated
service for delivering quality software products in distributed environments like
the Grids.

@@

The European Middleware Initiative ({\sl EMI}, May 2010 -- Apr 2013)
\cite{cordis:emi} project joined the 4 major Grid middleware providers in
Europe at the time -- {\sl gLite}, {\sl UNICORE}, {\sl ARC} and {\sl dCache} --
with the goal to maintain and evolve the middleware focusing on extending their
interoperability and improving the reliability of the services. The
ISO/IEC 9126 \cite{iso-9126} standard was used in order to identify a set of
characteristics that needed to be present in the {\sl EMI} software products and
processes to be able to meet the {\sl EMI} quality requirements
\cite{emi-quality-model}.

For each software characteristic, a set of associated
metrics and Key Performance Indicators (KPIs) were identified and defined in
detail in the {\sl EMI} Metrics Specification \cite{emi-metrics}. The project
leveraged the {\sl ETICS} service for the development, continuous integration and release management,
as well as for metric tracking, making queries on the collected data to display
them through the chart generation framework.

@@

The {\sl EGI--InSPIRE} (Integrated Sustainable Pan--European Infrastructure for
Researchers in Europe, May 2010 -- May 2014) project \cite{cordis:egi-inspire}
arose as the continuation of the {\sl EGEE--III} project, with the mission of establishing and
maintaining a sustainable European Grid Infrastructure.

{\sl EGI--InSPIRE} not only ensured the coordination of the operation of a secure, reliable 
European-wide production Grid infrastructure federated from National Grid Initiatives (NGIs)
that is integrated and interoperates with other Grids worldwide but also foreseed the 
maintenance and development of operational tools, like the Operations Portal \cite{egi-ops}, 
the {\sl EGI} Helpdesk \cite{ggus} or Grid configuration repository (GOCDB) \cite{gocdb} and the 
coordination of the external provision of software required by {\sl EGI} to 
form the production infrastructure. For the later it maintained a production--ready 
Grid computing middleware distribution called {\sl UMD} (Unified Middleware Distribution), and 
defined general and component specific quality criteria to be applied to software 
components made available through this repository.
The role of {\sl UMD} was to enforce the fulfilment of this set of quality
criteria \cite{egi-qc} in the software being delivered (i.e.,
{\sl EMI} and {\sl Globus}) \cite{mario}. {\sl UMD} is still being used and
deployed in the European scientific e--Infrastructures under a follow--up
project, the {\sl EGI--Engage} (Engaging the EGI Community towards an Open Science
Commons) \cite{cordis:egi-engage}. This distribution is currently complemented
by a Cloud--specific one called {\sl CMD} (Cloud Middleware Distribution).

@@

The {\sl INDIGO--DataCloud} (INtegrating Distributed data Infrastructures for Global
explOitation, Apr 2015 -- Sep 2017) \cite{cordis:indigo} is the last
project of software development considered in this review. At the time of writing
this paper, the project is succeeding in providing solutions to address the existing gaps in
Cloud Platform--as--a--Service (PaaS) and Software--as--a--Service (SaaS) levels,
helping developers, e--Infrastructures and scientific communities to exploit
Cloud computing benefits.

@@

{\sl INDIGO--DataCloud} is the most recent project included in this paper thus it
emerged from the lessons learned in the previous experiences described in
Section \ref{sec:ev}. The expertise gathered throughout these years was highly
profitable in terms of the evaluation and application of innovative software engineering
methodologies, the management of new technologies to put those methodologies
into practice, and the compilation and analysis of the appropriate set of metrics to measure the
success of the solutions implemented. All the above with the additional complexity of
being framed into large collaborative projects with an extensive list of partners.

@@

%The {\sl INDIGO--DataCloud} project reflects the progress made in software
%quality and reliability aspects throughout the past scientific European
%experiences. Nevertheless (I would remove this), the project’s Quality Assurance procedures are also
%highly influenced by the insights of current big worldwide collaborations of
%software development. These collaborations have continuously inspired the use of new
%technologies and procedures to increase the robustness and quality of the
%software being delivered. In this scenario, the implementation of the DevOps culture 
%is one of the major outcomes, as this paradigm has been progressively adopted by the
%project.

\begin{figure}[ht]
\centering
\includegraphics[width=0.5\textwidth]{images/codestyle.png}
\caption{Code style standards followed by {\sl INDIGO--DataCloud}'s software products.}
\label{fig:fig_codestyle}
\end{figure}

@@

\begin{itemize}
\item \textit{Code style}.
The ultimate goal for code style assessment is to improve the readability and reusability of the source code produced under the scope of the
project. Code style standards are enforced for every software component in the project stack. Community
most-adopted or \textit{de-facto standards} are initially recommended, however the development teams eventually
choose the set of guidelines to comply with.  Figure \ref{fig:fig_codestyle} highlights the code style standards and their popularity
over the {\sl INDIGO--DataCloud} software stack.

\item \textit{Unit and Functional testing}.
Changes involving the addition of new functionalities are requested to be tested. \textit{Regression
testing} in this context is done enforcing the successful execution of the previous
functionality tests already present in the code, so to uncover any malfunction introduced by
the changes. On the other hand, \textit{unit testing} completes the source code testing evaluation by
focusing on the code's internal design. The SQA criteria sets a recommended threshold of
70\% coverage for unit testing. In Figure \ref{fig:fig_unittest}, unit testing coverage is compared
for each software component over the two major {\sl INDIGO--DataCloud} releases, denoting an incremental 
trend over the course of the project. As depicted in the Figure, 50\% of the components were over the
70\% threshold while slightly less than 80\% of the product stack had values over 50\% coverage.

@@

\item \textit{Automated Deployment}.
In order to lower the barriers of software adoption, the SQA criteria requires the automated 
deployment of the products delivered as part of the catalogue. Automation in this context is 
tackled using a configuration management tool. These tools allow to have an 
\textit{infrastructure as code} (IaC), managing the component's deployment through declarative
definitions. The definition files appear as additional source of documentation, 
self-documenting code, as they sequentially guide the component's deployment process in multiple
platforms. In this regard, {\sl INDIGO--DataCloud} project contributed to open--source IaC tools such 
as Ansible \cite{indigo-ansible} and Puppet \cite{indigo-puppet}. A representative example of 
such contribution are the \textit{50 roles} being developed from scratch, currently hosted in
the Ansible Galaxy portal \footnote{https://galaxy.ansible.com/indigo-dc/}. Figure 
\ref{fig:fig_confman} shows the number of products that offer an automated means for deployment.
It shows an increase after the {\sl INDIGO--1} release especially in the previous weeks of each
of the two major releases.
\end{itemize}

@@

\subsubsection{Continuous Integration (CI)}
\label{subsec:ci}
The {\sl INDIGO--DataCloud} project promoted the application of a CI scenario that enforces,
for any piece of software produced, the testing requirements defined in the SQA criteria,
as described in Section \ref{subsec:sqa}. Such environment requires an automation
ready--to--go infrastructure where the different services and technologies involved interact with each
other to trigger the source code validation pipeline for each candidate change. The
pipeline is mainly comprised by the code style validation, unit and functional testing
coverage. The CI pipeline is complemented with additional quality checks, such as 
integration tests for the software components where automation of this type of testing
is applicable, or security linters for the code static analysis of suspicious
constructs that could lead to security risks. Metrics gathering also takes part of
the CI pipeline so to have a per-change trend evolution of the common source code 
related metrics, such as the source lines of code (SLOC) or the cyclomatic complexity.

@@

Figure \ref{fig:fig_jenkins_CI_builds} shows the evolution, throughout the 
{\sl INDIGO--DataCloud}'s first and second releases, of the required test types builds since the
implementation of the CI infrastructure. Automated functional testing coverage were not
available for all the software stack, thus the associated number of builds are fewer.

\subsubsection{Continuous Delivery}
As DevOps suggests, frequent releases positively affect the reliability of the software as
they involve fewer changes and usually more correlated. The software updates of 
{\sl INDIGO--DataCloud} products, taking place since the second major release, are passing through
a Continuous Delivery (CD) pipeline that adds the packaging of the software right after the
successful execution of the CI pipeline previously described. Consequently, CI-validated 
software components are automatically delivered in online repositories without the need of 
any human interaction but ultimately requiring expert supervision to validate the results.

{\sl INDIGO--DataCloud} software stack is delivered in form of Linux software packages (rpms, debs) 
and Docker containers. The CD approach is different for each type of software packaging. \textit{Linux
package pipeline} intentionally delivers the packages produced in a pre-production -- preview --
branch. Before that, once the software packages have been created, they are uploaded to a 
testing branch. The available automated deployment solution -- see \cite{subsec:sqa} -- checks
the installation and configuration of the component relying on the testing repository. On 
successful completion, the packages are then automatically signed and subsequently moved to 
the preview branch, from where they finally be made available in the production branch after
the release manager supervises the process.

The \textit{Docker container pipeline} uses instead the automated deployment solution to 
actually install and configure the software component in the container image. Once this is 
done the recently created image is uploaded to the production DockerHub repository, tagged
with the label corresponding to the current {\sl INDIGO--DataCloud} release. Figure 
\ref{fig:fig_CD} shows the complete workflow of the Docker container CD implementation, 
automatically triggered by a change in the source code.

@@

\begin{figure*}	
	\centering
	\begin{subfigure}
		\centering
\includegraphics[width=0.85\textwidth]{images/disvis-flow.png}
\caption{DevOps pipeline to distribute Docker images for Disvis application.}
\label{fig:fig_disvis}		
	\end{subfigure}
	\quad
	\begin{subfigure}
		\centering
\includegraphics[width=\textwidth]{images/pilotpreview.png}
\caption{Resource centers supporting the Pilot Preview testbed and corresponding
set of deployed {\sl INDIGO--DataCloud} components or services.}
\label{fig:fig_pilotpreview}
	\end{subfigure}
\end{figure*}

@@

The experience gathered throughout the project with regard to the adoption of
different DevOps practices is not only useful and suitable for the software related
to the core services in the {\sl INDIGO--DataCloud} solution, but also applicable to the
development and distribution of the applications coming from the user communities.

@@

User application developers were provided with both a means to validate the
source code before merging and the creation of a new versioned Docker image,
automatically available in the {\sl INDIGO--DataCloud}’s application repository.

@@

Two pilot infrastructures are at the disposal of developers and scientific
communities involved in the project. The aim of these testbeds is to test the
level of integration between the components involved in the {\sl INDIGO--DataCloud}
solutions and use cases validation, by deploying and executing the applications
with the last stable version of the software in environments as similar as 
possible to production ones. A map of the Pilot Preview
infrastructure is depicted in Figure \ref{fig:fig_pilotpreview}, it shows the
resources providers and the components or services they deployed and supported.
